Rapid Epidemiological Analysis of Comorbidities and Treatments in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population

### 2020-05-13 casecontrol.Rmd
multiple updates

table.densities <- rbind(summary(listed.densities),
                         summary(full.densities))[, 1:6]

@@

_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the odds ratios were

`r table.agegr.all[1, 3]` for residence in a care home status, 
`r table.agegr.all[4, 3]` for diabetes, 
`r table.agegr.all[8, 3]` for chronic renal failure, 
`r table.agegr.all[5, 3]` for ischemic heart disease
`r table.agegr.all[9, 3]` for neurological disease, 
`r table.agegr.all[10, 3]` for chronic liver disease and 
`r table.agegr.all[11, 3]` for immune deficiency or suppresssion

Overall but INSERT xx% of severe cases versus INSERT xx% of controls had at least one of these INSERT ( xx versus xx% in those aged under 40 years).  The odds ratio for severe disease associated with encashment of at least one prescription drug in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).  These associations with severe disease were present in those without any of the listed conditions, and were distributed over most BNF chapters and the associations with hospital diagnoses over most ICD-10 chapters.

In comparison with a baseline model based on care home residence only, adding listed conditions increased the information for discrimination to  INSERT 0.9 bits. Adding indicator variables for ICD-10 and BNF chapters increased the information for discrimination further to  INSERT 1.5 bits. From a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.

*Conclusions* --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history and the vast majority of younger persons developing severe disease have an underlying medical condition. The comorbidities associated with increased risk are not limited to the current lists of risk conditions highlighted by public health agencies.  A risk classifier that uses all the information available in medical and prescribing records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the SARS-CoV-2 epidemic is over.   The strong association with prescribing history we have observed may reflect co-morbidity or adverse drug effects and we are investigating this further.

@@

# Methods
## Case definition 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.  The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland.  All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.   Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and the weekly updated Scottish Morbidity Record 01. Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG)  database that captures admission to all critical care units across Scotland. Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")`
and cause of death were obtained from linkage to the National Register of Scotland.

We classified test-positive cases into three groups:

1. Severe or fatal COVID-19, defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test regardless of the cause of death given on the death certificate.

2. Hospitalised but not severe as defined above

3. Not hospitalised

By restricting the case definition to a group that combines those receiving critical care with those who died without receiving critical care, we ensure complete ascertainment of severe test-positive cases, whatever selection procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.

## Matched controls
For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case using the Community Health Index (CHI) database. Note that this is an incidence density sampling design i.e. a control who becomes a case at a later point is correctly retained in the control set. With this design, the odds ratios in a conditional logistic regression model are equivalent to rate ratios.

## Demographic data
Socioeconomic status was scored using the postcode based indicator  the Scottish Index of Multiple Deprivation using the postcode in the CHI database.  Ethnic group was captured in two ways: self designated ethnicity  was derived from past records in the hospital admissions and outpatient attendances  (Scottish Morbidity Records) using  categories based on those used in the national census. However since this is incomplete  we also used the a name classification algorithm ONOMAP [@lakha_name_2011] with the names in the CHI database.  This is able to distinguish South Asian and Chinese groups but not Black-Caribbean people from the White group.  Residential care home status was available via the CHI database.

## Morbidity and drug prescribing
For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record 01 (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge. British National Formulary  (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.  Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.

We first extracted history of a specific list of conditions and drug classes that have been  included as  risk conditions for COVID-19 by  public health agencies https://www.nhs.uk/conditions/coronavirus-covid-19/people-at-higher-risk-from-coronavirus/.  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.  The ICD-10 codes and drug codes used to define each variable are listed in Supplementary Materials.  In addition to these pre-specified groupings of codes we then undertook a systematic evaluation of the association of all ICD-10 codes  and BNF  codes to chapter level.  We did not  specifically  tabulate  the conditions that have been listed as making people " clinically extremely vulnerable"  such as cystic fibrosis or leukaemia  since there is no  clinical equipoise  regarding risk in such persons and since they will be underrepresented in the case data as they have been in complete shielding since early in the epidemic.

## Statistical methods
We calculated the sex and age band specific cumulative incidence of severe COVID-19 disease using the age and sex distribution of those alive in the population as of March 1st 2020. (CHECK)  The univariate association of  sociodemographic variables, pre-defined conditions and discharge and prescribing history  with severe COVID-19 were modelled with conditional logistic regression, implemented as Cox regression in the R function `survival::clogit` We examined the associations by age band and overall.  Among those  cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF  chapter with severe COVID-19. To construct a risk prediction models, we used stepwise regression alternating between forward and backward steps, implemented in the R function `stats::step`, so as to maximize the AIC.  The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic.

## Information governance
This research was conducted under approvals A(i) PBPP approvals for accessing ?????; (ii) existing PBPP approval that allows HPS staff to link any PHI controlled datasets; (janet to add detail and reference numbers?) (iii) PBPP amendment (1617-0147) that covers inward transfer of diabetes data. Datasets were de-identified before analysis.

# Results
<!---
Still needed

- A statement on "In national mortality data, the n and % of those certified with COVID-19 as underlying case of death had/ did not have a positive test record"  or sensivitivty analysis if sharon provides a case control set of the deaths with ue07 code 
- Figure - ?? graph of population attributable risk fraction of listed conditions by age ?? not sure if this is worth the trouble ...
-  Run table 2 analyses for all subjects combined 
- what is currently tables 3,4,5  and 7  could be put into appendix and the table reference in text changed as appropriate ?
- tables currently 9,11,12 should be removed as will go in pharmacoepi  paper  --->
## Incidence and mortality from severe COVID-19 in the Scottish population
Logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years.  Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.  The relationship of mortality to age was almost exactly linear on a logit scale.  For fatal and non-fatal cases combined, the mortality differentials between males and females widened in middle age before narrowing again after about age 75 years.  In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(logistic.coeffs[3, 2]), 2)` for fatal disease.  The odds ratio associated  with male sex was is INSERT  for all severe disease and  INSERT  for fatal disease.  As shown in Table INSERT  almost all of the severe cases  death over aged 75years died  but  the majority of severe cases under aged 60 survived.

@@

## Associations with listed conditions
Overall  INSERT xx of cases and xx of controls had at least one of the listed conditions.   The  prevalence of the listed conditions in cases and controls by age band is shown in Table  INSERT .   Almost half of cases aged under 40 years had at least one listed condition, compared with only 9% (43 / 470) of controls.  In those 75+ years INSERT xxx 80% of cases had at least one listed condition but  INSERT xx% of controls also did.

As shown  in the age band specific univariate logistic regression models in tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75}.  Here we collapsed the two younger age bands to <60 years.  All of the listed conditions were significantly associated with severe disease  in each age band. Diabetes was associated with a 6.6. fold rate of severe COVID-19 under age 60 years.  Of note is that "other heart disease" was more  strongly associated with severe COVID-19 than ischaemic heart disease.

@@

## Systematic analysis of diagnoses associated with severe disease
The approximately two-fold  risk of  with severe  COVID-19  associated with prior hospital admission independently  of other conditions and factors was further explored. The association of hospital diagnoses of each ICD-10 chapter with severe COVID-19, restricted  to those without any of the listed conditions.  These results are shown in Table \ref{tab:conditions}.  In univariate analyses, almost all ICD chapters with at least one diagnosis, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  In a multivariate analysis of all chapters, the most strongly associated chapter-level diagnosis was Chapter 2 (neoplasms).  This was further disaggregated with subchapters in as shown in Supplement  Table \ref{tab:icdchapter2}. Unexpectedly, the strongest association was with the subchapter "Malignant Neoplasms of Digestive Organs".  Several other subchapters corresponding to cancer sites were also associated with severe disease.

## Associations of prescribed drugs with severe disease 
Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of any prescription in the last year.  The rate ratio associated with this variable varies from INSERT 9.2 in those aged under 60 years to INSERT 42.2 in those aged 75 years and over.  In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to  INSERT3.8 and INSERT11.9 respectively.  It is notable that about one third of those over 75 had not encashed a prescription in the previous year.

@@

For each of variable set, a stepwise regression procedure was carried out using alternating forward-backward selection. The variables retained with each variable set are shown in Table \ref{tab:stepwise} . The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation.  Observed and predicted case status were compared over all test folds.  Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[1, 3]` to `r table.densities[2, 3]` and the expected information for discrimination from `r table.densities[1, 5]` bits to `r table.densities[2, 5] ` bits.

This value for the information conditional on age and sex can be added to the information for discrimination obtained with a logistic regression model using age and sex to obtain the total information for discrimination that would be obtained in the population  to give a total of WHAT??.  PAUL HOW MUCH IS THAT THIS IS NOT CLEAR

Figure \ref{fig:wevid} shows the class-conditional distributions of the weight of evidence favouring case over <!--thi figure should go into a supplement--> control status from the model based on the extended variable set.  This shows, as expected for a multifactorial classifier, that the distributions are approximately Gaussian: there is no clear divide between high-risk and low-risk individuals of the same age and sex.

@@

@@

knitr::kable(table.demog.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="demog",
             digits=c(0, 0, 0, 200), 
             col.names=c(colnames(table.demog.aug)[1:2],
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Univariate associations of severe disease with demographic factors") %>%
    column_spec(column=1, width="3cm") %>%
    group_rows("Ethnicity", 1, 3) %>%
    group_rows("SIMD quintile", 4, 8)

@@

```
```{r listed.condition.75plus, echo=FALSE, warning=FALSE, message=FALSE}

@@

```{r bnfchapter1, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.bnfchapter1,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter1",
             digits=c(0, 0, 0, 200 , 0, 200), 
             col.names=c(colnames(table.bnfchapter1)[1:2],
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with prescribed drugs in BNF chapter 1, in those without listed conditions") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    landscape()
```

```{r protonpump, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.protonpump,
             escape=FALSE, 
             booktabs=TRUE,
             label="protonpump",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(colnames(table.drugs.aug)[1:2],
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Univariate associations of severe disease with proton pump inhibitors in all cases and controls, by age group") %>%
    column_spec(column=1, width="2cm")
```

```{r bnfchapter2, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.bnfchapter2,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter2",
             col.names=c(colnames(table.bnfchapter2)[1:2],
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with prescribed drugs in BNF chapter 2, in those without listed conditions") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    group_rows("", 1, nrow(table.bnfchapter2)) %>%
    landscape()

```{r bnfchapter4, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.bnfchapter4,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter4",
             col.names=c(colnames(table.bnfchapter4)[1:2],
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with prescribed drugs in BNF chapter 9, in those without listed conditions") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    group_rows("", 1, nrow(table.bnfchapter4)) %>%
    landscape()

@@

rownames(table.densities) <- c("Care home + listed conditions",
                               "Extended variable set")

@@

@@

stepwise.full <- as.data.frame(stepwise.full[, c(1, 5)])
stepwise.full[, 1] <- round(stepwise.full[, 1], 2)
stepwise.full[, 2] <- signif(stepwise.full[, 2], 1)
rownames(stepwise.full) <- replace.names(rownames(stepwise.full))
rownames(stepwise.full) <- gsub("\\_", " ", rownames(stepwise.full))

stepwise.full[, 2] <- pvalue.latex(stepwise.full[, 2])

knitr::kable(stepwise.full, 
             booktabs=TRUE,
             escape=FALSE,
             col.names=c("log rate ratio", "p-value"), 
             label="stepwise",
             align=rep("c", 2),
             caption="Stepwise regression: variables retained in model for severe disease") %>%
    group_rows("Diabetes", 3, 5)

@@

<!--

For each of these chapters, we examined univariate and multivariate associations with severe disease at the level of paragraph or subparagraph of the BNF classification, restricting the analysis to those without one of the listed conditions.

Table \ref{tab:bnfchapter1} shows that prescriptions of most classes of drugs for gastrointestinal conditions were associated with severe COVID-19.  By far the most widely prescribed drug class in this age group was proton pump inhibitors; 15% of controls and 36% of cases had encashed at least one prescription, with a univariate rate ratio of 3.2.  
In a multivariate analysis, the drug class most strongly associated with severe disease was proton pump inhibitors.

Table \ref{tab:protonpump} shows the association of severe disease with proton pump inhibitors in all cases and controls by age group. The univariate rate ratio varied from 12 in those aged under 60 years to 2.8 in those aged 75 years and over.

Table \ref{tab:bnfchapter2} shows the associations with classes of drugs prescribed for the cardiovascular system.  Most classes of antihypertensive drug were associated with increased risks in univariate analysis, but there was no evidence that the risks associated with prescription of ACE inhibitors or angiotensin-II receptor antagonists differed from those for other drugs.  Prescription of anticoagulants was associated with increased risk of severe disease. presumably reflecting the risks associated with the underlying cardiovascular conditions. In a multivariate analysis the strongest association of severe disease was with loop diuretics.  Table \ref{tab:bnfchapter4} shows the associations with classes of drugs prescribed for the central nervous system.  The strongest association with severe disease in a multivariate analysis was with prescription of non-opiate analgesics.

A striking finding of this study was the very strong association of severe COVID-19 with having encashed at least one prescription in the past year. This association was only partly explained by associations with listed conditions.  Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19.

Examination of the subchapters within this chapter showed that this association was largely explained by prescriptions of proton pump inhibitors, which had been encashed by 15% of controls.  Over all age groups the rate ratio associated with use of proton pump inhibitors was 2.8, but in those aged less than 40 years, in whom serious underlying disease is presumaby less common, the rate ratio was 12. This association is not explained by association any underlying gastrointestinal disease that leads to hospital admission, by coprescribing with dual antiplatelet agents in people who have had a myocardial infarction or a vascular procedure.  As proton pump inhibitors are available over the counter, our study of encashed prescriptions does not capture all exposure to this class of drugs.  However as there are no prescription charges in Scotland, long-term users of these expensive drugs are likely to seek to obtain them on prescription.

As proton pump inhibitors are commonly co-prescribed with NSAIDS, this is relevant to evaluating possible explanations for the association of proton pump inhibitors with NSAIDs. (CHECK).

As proton pump inhibitors are associated with increased risk of enteric infections, it is plausible that they could increase the risk of severe infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].  Others have noted that proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@marks_time_2016].  35% of patients admitted to acute geriatric units in Aberdeen had been inappropriately prescribed proton pump inhibitors: overprescribing was associated with the number of other drugs prescribed and with an index of comorbidity [@jarchow-macdonald_prescribing_2013].

-->

>>>>

table.densities <- rbind(summary(demog.densities),
                         summary(listed.densities),
                         summary(full.densities))[, 1:6]

@@

_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the odds ratios were 
`r table.agegr.all[1, 3]` for residence in a care home status, 
`r table.agegr.all[4, 3]` for diabetes, 
`r table.agegr.all[8, 3]` for chronic renal failure, 
`r table.agegr.all[5, 3]` for ischemic heart disease
`r table.agegr.all[9, 3]` for neurological disease, 
`r table.agegr.all[10, 3]` for chronic liver disease and 
`r table.agegr.all[11, 3]` for immune deficiency or suppresssion

Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 2]` of controls had at least one of these listed conditions: `r gsub("\\\\", "", table.agegr[13, 2])` versus `r gsub("\\\\", "", table.agegr[13, 1])` in those aged under 40 years).  The odds ratio for severe disease associated with encashment of at least one prescription drug in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).  These associations with severe disease were present in those without any of the listed conditions, and were distributed over most BNF chapters and the associations with hospital diagnoses over most ICD-10 chapters.

With a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.  The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, `r table.densities[2, 5]` bits for a model that included listed condtions, and `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.

_Conclusions_ --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history and most younger persons developing severe disease have an underlying medical condition. The comorbidities associated with increased risk are not limited to the current lists of risk conditions highlighted by public health agencies.  A risk classifier that uses all the information available in medical and prescribing records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the SARS-CoV-2 epidemic is over.   The strong association with prescribing history we have observed may reflect co-morbidity or adverse drug effects and we are investigating this further.

@@

# Methods
## Case definition 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.  The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland.  All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.  Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and the weekly updated Scottish Morbidity Record 01. Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care units across Scotland. Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.

Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  By restricting the case definition to a group that combines those receiving critical care with those who died without receiving critical care, we ensured complete ascertainment of severe test-positive cases, whatever selection procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.

## Matched controls
For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case using the Community Health Index (CHI) database. As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.  With this design, odds ratios in conditional logistic regression models are equivalent to rate ratios.

## Demographic data
Socioeconomic status was scored in the CHI database as the Scottish Index of Multiple Deprivation, an indicator based on postal code.  Ethnicity was assigned based on applying a name classification algorithm ONOMAP [@lakha_name_2011] to the names in the CHI database.  This classification is able to distinguish South Asian and Chinese groups with fair accuracy but does not distinguish those who identify as Black or Black-Caribbean from those who identify as White. For those cases and controls who had a past record of hospital admission or outpatient attendance in the Scottish Morbidity Record, a separate classification of ethnicity was available, assigned by self-report or by hospital staff and based on the categories used in the Census.

## Morbidity and drug prescribing
For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge. British National Formulary  (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.  Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.

We first extracted history of a specific list of conditions and drug classes that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.  The ICD-10 codes and drug codes used to define each variable are listed in the R script provided as Supplementary Material. In addition to these pre-specified groupings of codes we then undertook a systematic evaluation of the association of all ICD-10 codes and BNF codes to chapter level.  We did not specifically tabulate the conditions that have been listed as making people " clinically extremely vulnerable" such as cystic fibrosis or leukaemia as there is no clinical equipoise regarding risk in such persons and as it is expected that they will be underrepresented in the case data as they have been advised to shield themselves completely since early in the epidemic.

## Statistical methods
To estimate the relation of cumulative incidence and mortality from COVID-19 to age and sex, logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years. For the case-control study, estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.  Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19. To construct risk prediction models, we used stepwise regression alternating between forward and backward steps to maximize the AIC, implemented in the R function `stats::step`.  The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic.

## Information governance
This study was conducted under approvals from the Public Benefit and Privacy Panel that allow Health Protection Scotland staff to link any PHI controlled datasets <!--- janet to add detail and reference numbers ---> and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland. Datasets were de-identified before analysis.

# Results
<!---
Still needed 
- A statement on "In national mortality data, the n and % of those certified with COVID-19 as underlying case of death had/ did not have a positive test record"  or sensivitivty analysis if sharon provides a case control set of the deaths with ue07 code 
- Figure - ?? graph of population attributable risk fraction of listed conditions by age ?? not sure if this is worth the trouble ...
--->
## Incidence and mortality from severe COVID-19 in the Scottish population
Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.  The relationship of mortality to age was almost exactly linear on a logit scale, and the lines for males and females were almost exactly parallel.  For severe cases as defined in this study, the incidence appears to rise more slowly from age 50 to 70 years in females than in males, implying that the ratio of non-fatal cases who receive critical care to fatal cases is lower in females than in males in this age range.  This may reflect either an underlying divergence in case fatality, or that women are less likely than men to be transferred to critical care if they have a good chance of surviving without critical care.  In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(logistic.coeffs[3, 2]), 2)` for fatal disease.  The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.  As shown in Table \ref{tab:agegr} almost all severe cases over aged 75 years died but most severe cases under aged 60 survived.

@@

## Associations with listed conditions
Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of the listed conditions.  The prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.  `r gsub("\\\\", "", table.agegr[13, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[13, 1])` of the controls.  In those aged 75+ years `r gsub("\\\\", "", table.agegr[13, 6])` of the cases and `r gsub("\\\\", "", table.agegr[13, 5])` of the controls had at least one listed condition.

As shown  in the age band specific univariate logistic regression models in tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75}.  Here we collapsed the two younger age bands to <60 years.  All of the listed conditions were significantly associated with severe disease  in each age band. Diabetes was associated with a 6.6 fold rate of severe COVID-19 under age 60 years.  Of note is that "other heart disease" was more  strongly associated with severe COVID-19 than ischaemic heart disease.

@@

## Systematic analysis of diagnoses associated with severe disease
The approximately two-fold risk of with severe COVID-19 associated with prior hospital admission independently of other conditions and factors was further explored. The association of hospital diagnoses of each ICD-10 chapter with severe COVID-19, restricted to those without any of the listed conditions.  These results are shown in Table \ref{tab:conditions}.  In univariate analyses, almost all ICD chapters with at least one diagnosis, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  Table \ref{tab:icdsubchapters} extracts univariate associations with ICD subchapters: .. note associations with previous pneumonia and stroke.

## Associations of prescribed drugs with severe disease 
Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of any prescription in the last year.  The rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.  In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and  `r tables.agegr[[3]][2, 5]`   About one third of those over 75 had not encashed a prescription in the previous year.

@@

For each of variable set, a stepwise regression procedure was carried out using alternating forward-backward selection. The variables retained with each variable set are shown in Table \ref{tab:stepwise} . The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation.  Observed and predicted case status were compared over all test folds.  Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[2, 3]` to `r table.densities[3, 3]` and the expected information for discrimination from `r table.densities[2, 5]` bits to `r table.densities[3, 5] ` bits.

This estimate of `r table.densities[3, 5] ` bits for the information conditional on age and sex obtained from the matched case-control study can be added to the information for discrimination `r round(cases.Lambda.agesex, 2)` bits obtained from the logistic regression on age and sex in the population using age and sex to estimate the total information for discrimination of a risk classifier in the population as that would be obtained in the population to give a total of `r table.densities[3, 5] + round(cases.Lambda.agesex, 2)` bits.

Figure \ref{fig:wevid} shows the class-conditional distributions of the weight of evidence $W$ favouring case over control status from the model based on the extended variable set.  This shows, as expected for a multifactorial classifier, that the distributions are approximately Gaussian: there is no clear divide between high-risk and low-risk individuals of the same age and sex. Figure \ref{fig:roc}, which shows the corresponding receiver operating characteristic (ROC) curve, encodes exactly the same information in a different form.  The likelihood ratio $2^W$ at any given sensitivity and specificity is the gradient of the model-based ROC curve.

@@

```{r plotroc, echo=FALSE, fig.cap="\\label{fig:roc}Cross-validation of model chosen by stepwise regression using extended variable set: receiver operating characteristic curve", out.width="0.8\\textwidth"}

plotroc(full.densities) +
    scale_color_manual(values=c("orange", "purple")) + 
    coord_fixed(ratio=1) +
    theme(axis.title=element_text(size=18)) +
    theme(legend.title=element_blank()) + 
    theme(legend.position=c(0.8, 0.2)) + 
    scale_x_reverse(limits=c(1, 0), expand=c(0, 0)) +
    scale_y_continuous(limits=c(0, 1), expand=c(0, 0)) +
    xlab("Specificity") +
    ylab("Sensitivity (reverse scale)")

@@

knitr::kable(table.demog.aug[, 1:4],
             escape=FALSE, 
             booktabs=TRUE,
             label="demog",
             digits=c(0, 0, 0, 200), 
             col.names=c(colnames(table.demog.aug)[1:2],
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Univariate associations of severe disease with demographic factors") %>%
    column_spec(column=1, width="2.5cm") %>%
    column_spec(column=2:4, width="2.5cm") %>%
    group_rows("Ethnicity", 1, 3) %>%
    group_rows("SIMD quintile", 4, 8)

@@

@@

@@

rownames(table.densities) <- c("Demographic only",
                               "Demographic + listed conditions",
                               "Extended variable set")

@@

```{r listed.condition.75plus, echo=FALSE, warning=FALSE, message=FALSE}

```{r icdsubchapters, echo=FALSE, warning=FALSE, message=FALSE}
clean.header <- function(x) {
    x <- gsub("\\.\\.", " (", x)
    x <- gsub("\\.",  ")", x)
    return(x)
}

knitr::kable(table.icdsubchapters,
             escape=FALSE, 
             booktabs=TRUE,
             label="icdsubchapters",
             col.names=c(clean.header(colnames(table.icdsubchapters)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with hospital diagnoses by ICD subchapters: rows retained are those with p < 0.001 and at least 20 cases and controls") %>%
    column_spec(column=1, width="13cm") %>%
    column_spec(column=2:3, width="1.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    landscape()

@@

retained.full <- as.data.frame(stepwise.full[, c(1, 5)])
retained.full[, 1] <- round(retained.full[, 1], 2)
retained.full[, 2] <- signif(retained.full[, 2], 1)
rownames(retained.full) <- replace.names(rownames(retained.full))
rownames(retained.full) <- gsub("\\_", " ", rownames(retained.full))

retained.full[, 2] <- pvalue.latex(retained.full[, 2])

knitr::kable(retained.full, 
             booktabs=TRUE,
             escape=FALSE,
             label="stepwise",
             col.names=c("log rate ratio", "p-value"), 
             align=rep("c", 2),
             caption="Stepwise regression: variables retained in model for severe disease") %>%
    group_rows("Diabetes", 6, 8)

@@
